---
title: "DSC 520 Project"
author: "Nathan Reed"
date: "2023-05-14"
output: html_document
---

# Step 1

## Introduction
I want to research the topic of how to find students that are at risk of not 
graduating College. This is important because there are lots of students 
who drop out of school each year due to many factors. This research would help administrators be able to find students that are more likely to drop out and 
help prevent them from dropping out. This is a data science problem because Data Scientists are able to use data to find patterns that would help them identify 
these at risk students more likely.

## Research Questions
1. Is there a correlation between demographic and drop outs?  
2. Does poor academic performance or work ethic contribute more to dropping out?  
3. Are students with disabilities more likely to drop out than neurotypical students?  
4. If the parents dropped out does that mean that student is more likely to drop out?  
5. Does socioeconomic status effect the likelihood of dropping out?  
6. Is where you live significant relating to the state you are studying in?  
7. Is gender significant in the effect of dropping out?  

## Approach
I will use data from different sources to answer the questions that I can with 
each individual dataset. I will then add all the key variables that answer the 
questions above. I will then make sure all linearity, independent, normality, 
and equal variance assumptions are correct. I would make sure that it is clean 
and then run any transformations that are needed. I would then use anova to 
decide the model of best fit based off of adjusted $r^2$. I would then use plots
and tables to decide what is significant and not for the research questions 
above.

## How your approach addresses the problem
The approach mentioned should help answer the overarching problem I want to 
solve as well as the research questions because of the significance tests that 
I will use to prove if they effect drop out rates or not.

## Data
Valentim Realinho, Jorge Machado, Luís Baptista, & Mónica V. Martins. (2021). Predict students' dropout and academic success (1.0) [Data set]. Zenodo. https://doi.org/10.5281/zenodo.5777340

This data was originally used for predicting student dropouts and academic 
success. 24 different variables. 

https://data.world/databeats
The data is sourced from two primary sources – the National Center for Education Statistics (NCES)’ Integrated Postsecondary Education System (IPEDS) and 
Voluntary System of Accountability’s Student Success and Progress rate. 

Martínez, Alvaro; Verdú, Elena; Moreno-Ger, Pablo (2019), “University dropout experiment”, Mendeley Data, V1, doi: 10.17632/pn8k5xp37c.1

## Required Packages
```{r}
library(dplyr)
library(ggplot2)
library(corrplot)
library(car)
library(purrr)
```

## Plots and Tables
I would make sure to do qqplots, scatterplots, histograms, boxplot, correlation 
matrix, summary statistics, and anova.

## Questions for Future Steps
Should I consolidate all my data for this or use 3 individual datasets to answer
different questions?
How to make a model with more than one dataset being used?

# Step 2

## How to import and clean my data
I first read in the data from each different dataset.
```{r}
setwd("C:/Users/ncree/OneDrive/Documents/DSC 520")
dropout_df1 <- read.csv("dropout_data_1.csv")
dropout_df2 <- read.csv("dropout_data_2.csv")
dropout_df3 <- read.csv("dropout_data_3.csv")
```
After reading in all the datasets, I first made sure there was no duplicates. I 
Then made sure to keep only the variables that I want to use for the data. I 
then made sure that all the variable names work so that I can use them easy. I 
then made all of the NA's in grad_rate on the dropout_df2 dataframe a 0.0 so we 
can use it in the graphs.
```{r}
# finds any duplicates in the data
dropout_df1 <- dropout_df1[!duplicated(dropout_df1), ]
dropout_df2 <- dropout_df2[!duplicated(dropout_df2), ]
dropout_df3 <- dropout_df3[!duplicated(dropout_df3), ]
# keeps only columns that I want for each data set
drop1 <- c("Application.mode","Application.order", "Course", "International",
           "Daytime.evening.attendance", "Age.at.enrollment",
           "Curricular.units.1st.sem..without.evaluations.",
           "Curricular.units.2nd.sem..without.evaluations.",
           "Curricular.units.2nd.sem..approved.", 
           "Curricular.units.1st.sem..approved.", 
           "Curricular.units.2nd.sem..credited.", 
           "Curricular.units.1st.sem..credited.", 
           "Curricular.units.2nd.sem..enrolled.", 
           "Curricular.units.1st.sem..enrolled.",
           "Curricular.units.1st.sem..evaluations.",
           "Curricular.units.2nd.sem..evaluations.")
dropout_df1 <- dropout_df1[,!(names(dropout_df1) %in% drop1)]
drop2 <- c("index","stateid", "state_abbr", "level", "year", "cohort", 
           "grad_cohort_ct", "control")
dropout_df2 <- dropout_df2[,!(names(dropout_df2) %in% drop2)]
drop3 <- c("Vulnerable_group", "Desired_Program", "H_subjects", "STEM_subjects")
dropout_df3 <- dropout_df3[,!(names(dropout_df3) %in% drop3)]
# changing variable names
colnames(dropout_df1)[c(2:8)] <- c("Prev.Education","Prev.Education.Grade", 
                                  "Nationality", "Mom.Education", 
                                  "Dad.Education", "Mom.job", "Dad.job")
colnames(dropout_df1)[c(16,17)] <- c("First.Sem.Grade", "Second.Sem.Grade")
# changing all NA's to a 0.0 for the percentage in grad rate
dropout_df2[is.na(dropout_df2)] <- 0.0
dropout_df2$dropout <- dropout_df2$grad_cohort - (dropout_df2$grad_100 +
                                                    dropout_df2$grad_150)
attach(dropout_df2)
dropout_df2 <- dropout_df2[order(state),]
dropout_df1$Dropout[dropout_df1$Target == "Dropout"] <- 1
dropout_df1$Dropout[dropout_df1$Target == "Graduate"] <- 0
dropout_df3$Dropout[dropout_df3$Dropout == "YES"] <- 1
dropout_df3$Dropout[dropout_df3$Dropout == "NO"] <- 0
```


## What does the final data set look like?
```{r}
head(dropout_df1)
head(dropout_df2)
head(dropout_df3)
```

## Questions for future steps
I feel like I was able to import and do everything that I wanted to do to clean 
the data. The only thing that I may want to learn is how to combine all the 
imported data.

## What information is not self-evident?
On the second data set, there is not really a dropout column for the dataset. I 
will need to create is using the other columns for time that take over 100% and 
150% of the normal time it takes to graduate.

## What are different ways you could look at this data?
I believe I will be able to answer all the questions except for substance abuse 
by just comparing the variable for the category to those who dropped out. For 
substance abuse I will probably need to find another dataset or make some 
assumptions about those who are more likely to abuse substances.

## How do you plan to slice and dice the data?
I do plan on creating more columns for the data by using other columns in the 
data. An example of this is mentioned above, but I would like to create a column 
for dropout for the second dataframe. Other than that, the data seems clean and 
ready enough for modeling.

## How could you summarize your data to answer key questions?
I would create a model and run anova tests to see if one group is more likely to
dropout. I then would also use the summary function to see if any of the 
variables is significant.

## What types of plots and tables will help you to illustrate the findings to your questions?
There isn't really a table that will help illustrate my findings for my 
questions. I would use Anova and summary tables in order to answer the questions 
that I had. I would use it for finding the significant variables as well as the 
student who is most likely to dropout.

A plot that would help with the data is histograms. I am mainly working with 
categorical variables, so I can run each variable and see which ones are more 
likely to produce dropouts.
```{r}
dropout_lm1 <- lm(Dropout~ Prev.Education.Grade + Educational.special.needs, 
                  data = dropout_df1)
summary(dropout_lm1)
ggplot(dropout_df2,aes(x = race, y = dropout)) + geom_bar(stat = "identity") +
  scale_y_continuous(expand = c(0,0), limits = c(0,40000000))
ggplot(dropout_df2,aes(x = gender, y = dropout)) + geom_bar(stat = "identity") +
  scale_y_continuous(expand = c(0,0), limits = c(0,40000000))
ggplot(dropout_df2[1:21235,],aes(x = state, y = dropout)) + 
  geom_bar(stat = "identity") + scale_y_continuous(expand = c(0,0), 
                                                   limits = c(0,40000000))
ggplot(dropout_df2[21236:42471,],aes(x = state, y = dropout)) + 
  geom_bar(stat = "identity") + scale_y_continuous(expand = c(0,0), 
                                                   limits = c(0,40000000))
ggplot(dropout_df2[42472:63706,],aes(x = state, y = dropout)) + 
  geom_bar(stat = "identity") + scale_y_continuous(expand = c(0,0), 
                                                   limits = c(0,40000000))
ggplot(dropout_df2[63707:84942,],aes(x = state, y = dropout)) + 
  geom_bar(stat = "identity") + scale_y_continuous(expand = c(0,0), 
                                                   limits = c(0,40000000))
dropout_lm2 <- lm(dropout~ race + gender + state, data = dropout_df2)
summary(dropout_lm2)
ggplot(dropout_df3,aes(x = Dropout, y = Family_income)) + 
  geom_bar(stat = "identity")
socioeconomic <- table(dropout_df3$Socioeconomic_level, dropout_df3$Dropout)
mosaicplot(socioeconomic)
Fatherlevel <- table(dropout_df3$Father_level, dropout_df3$Dropout)
mosaicplot(Fatherlevel)
Motherlevel <- table(dropout_df3$Mother_level, dropout_df3$Dropout)
mosaicplot(Motherlevel)
dropout_lm3 <- lm(Dropout~ Socioeconomic_level + Family_income + Father_level +
                    Mother_level, data = dropout_df3)
summary(dropout_lm3)
```

As seen from this graph we can tell that it seems that there are a lot more 
Caucasians who drop out then others. We can also tell from the states that the
states with more population like Texas, New York, and California have more 
dropouts. Then with gender ones with both are more likely to drop out than both 
male and female. Both the state and demographic question do not explain the 
dropout rates as this could easily be connected that there are more people in 
those states than others as well as there might be more Caucasians at college 
than other demographics.

## Do you plan on incorporating any machine learning techniques to answer your research questions? Explain.
I do not plan on using it machine learning to predict what type of students are 
more likely to drop out. Other than that, I do not plan on using it as this does
not need much machine learning to work.

## Questions for future steps.
How to best use predictions for this dataset in order to predict what type of 
student is most likely to dropout of college?

# Step 3

## Introduction
I have been able to learn through this data that there are groups that are more
likely to dropout from college than others. As mentioned in previous sections I 
would still need to make sure that the data does not have causation that could 
be affecting the results. I would need to make sure that there is no other 
reason for the type of student to dropout more than others.  

## The problem statement you addressed
I addressed the problem of what type of student is more likely to dropout of 
college than others. This is an important issue due to the fact that there are a 
lot of people who dropout of school early and are unable to find jobs. With more 
jobs requiring more education, this is a serious issue and it could be used to 
find out whether high school or even middle school students are more likely to
dropout.

## How you addressed this problem statement
I looked at the gender, demographic, previous generation education, the state 
they learned in, socioeconomic status, work ethic, or if they had a learning 
disability. I used summary statistics and boxplots to see if one group made it 
more likely to dropout.

## Analysis
I was able to find out that people consider themselves both genders are more 
likely to dropout than both male and female. I was also able to find out that 
people who either don't disclose their race or who did not fall under the 
categories listed are more likely to dropout. I was also able to note that 
Texas, California, and New York have more dropouts, however this could be more 
related to the population size than they are more likely to have dropouts. I was 
also able to find out that previous education grade was significant in how well 
students performed while if you had disabilities it was not. We can also tell
that, while not a huge indicator, if your family has more income you are less 
likely to dropout. I was also able to find that the parents level of education
was not as helpful as I would have thought. neither were significant and you 
could barely tell the difference in the mosaic graph that I did. 

## Implications
The implication of this data is that there are some student who will fall in 
this category that would not end up dropping out but because they are more 
likely to dropout, they would be forced to get help that they do not really 
need. Another big issue would be that people that fall into this category would
be more likely to give up and dropout because they know that they are more 
likely to dropout anyways. There is also the issue that finding ways to help 
these students could be expensive and not all schools have the budget to offer 
programs to help.

## Limitations
While this is real data, it is not perfect as there are many data points that 
are not useful. One example of this is for state there were way more dropouts in
a state called the United States. This is not a state obviously, but they were 
unable to get the exact state. This data is not very useful for what states are
more likely to have dropouts. Another issue is that I was unable to come up with
an effective way to compare groups that are the majority compared to the 
minority. Another limitation I had was how to make all the data I had combined
so I could make an effective model.

## Concluding Remarks
I thought this project helped me realize there is still a lot to learn how to 
do. I was able to get some results that I was looking for that made the project 
very interesting. I know that there is still a lot to learn with how to make 
data fit together, make sense, and be able to be used to make analysis. 
